---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# EPAC model

<!-- badges: start -->
<!-- badges: end -->

This package implements the Early Pandemic Age-Structured Compartmental (EPAC) model developed by [@wzmli](https://github.com/wzmli) and [@papsti](https://github.com/papsti) at [@phac-nml-phrsd](https://github.com/phac-nml-phrsd) using  [`macpan2`](https://github.com/canmod/macpan2) modelling software. **This package is still in development.**

The goal of this package is to document the models being used under this framework. It is more a software record rather than a package to help one run the model. 

## Installation

You can install the development version of EPACmodel like so:

``` r
# FILL THIS IN! HOW CAN PEOPLE INSTALL YOUR DEV PACKAGE?
```

## Getting started

```{r}
library(EPACmodel)
```

### Set up model simulator

To work with a model, we need to load its simulator. A simulator in `macpan2` is an object that includes model structure (state names, flows, etc.) along with parameter values and initial states. 

This package includes several models whose simulators can quickly and easily be retrieved. Available models include:

```{r}
list_models()
```

To get this model's simulator, we simply call:

```{r}
model_name <- "two-age-groups"
model_simulator <- make_simulator(
  model_name = model_name
)
```

By default, `make_simulator()` will attach a set of default parameters and initial states to the model structure. It will also set a default number of time steps. Any of these can be changed by passing additional, optional, arguments to `make_simulator()`. We recommend first loading the default parameters and/or states, and then editing the values in these lists in-session, to ensure the format of each object is compatible with the model definition and `macpan2`:

```{r}
default_params = get_params(model_name)
print("default params:")
print(default_params)

default_state = get_state(model_name)
print("default state:")
print(default_state)

# move some young susceptibles to the recovered class
new_state = default_state
new_state["R_y"] = 1000
new_state["S_y"] = default_state["S_y"] - new_state["R_y"]
print("new state:")
print(new_state)
```

Then you can pass the modified params and/or state to `make_simulator()`:

```{r}
new_model_simulator = make_simulator(
  model_name = model_name,
  state = new_state,
  time_steps = 200
)
```

### Simulate base model

To run the simulation:

```{r}
sim_output = simulate(model_simulator)
```

This output will include the value of each state at the given time (`value_type == 'state')`:

```{r}
(sim_output
 |> dplyr::filter(value_type == 'state', time == 10)
 |> head()
)
```

as well as the inflow into each compartment at a given time (`value_type == 'total_inflow')`, so that one could look at incidence, for instance:

```{r}
(sim_output
 |> dplyr::filter(value_type == 'total_inflow', time == 10)
 |> head()
)
```

We can plot the results using standard data manipulation and plotting tools:

```{r}
plot_output <- function(output){
  (output
 # parse state names
 |> dplyr::mutate(
   epi_state = gsub("_(y|o)", "", state_name),
   age = ifelse(grepl("_y", state_name),"young","old")
 )
 |> dplyr::filter(value_type == 'total_inflow', epi_state == 'I')
 |> ggplot2::ggplot()
 + ggplot2::geom_line(ggplot2::aes(x = time, y = value, colour = age),
                      linewidth = 1.25)
 + ggplot2::scale_y_continuous(
   labels = scales::label_number(scale_cut = scales::cut_short_scale())
 )
 + ggplot2::labs(
   x = "day",
   title = "Incidence over time by age group"
 )
 + ggplot2::theme_bw()
 + ggplot2::theme(
   axis.title.y = ggplot2::element_blank(),
   legend.position = c(1,1),
   legend.justification = c(1,1),
   legend.background = ggplot2::element_rect(fill = NA)
  )
)
}
```

```{r sim-output-1}
plot_output(sim_output)
```


### Model with interventions

Here is the base model with interventions to reduce the transmission rate twice over the course of the simulation:
  
```{r, eval = FALSE}
# simulate and create base plot
p <- ("two-age-groups_interventions"
  |> make_simulator()
  |> simulate()
  |> plot_output()
)

# annotate base plot with intervention timing
interventions <- data.frame(
  time = c(40, 50),
  y = 2.5e5,
  label = "intervention implemented"
)

(p
  + geom_vline(
    data = interventions,
    mapping = aes(xintercept = time),
    linetype = "dashed"
  )
  + geom_text(
    data = interventions,
    mapping = aes(x = time - 1, y = y, label = label),
    angle = 90,
    vjust = "inward"
  )
)
```

