---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  # eval = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "50%"
)
```

# EPAC model

<!-- badges: start -->
<!-- badges: end -->

This package implements the Early Pandemic Age-Structured Compartmental (EPAC) model developed by [@wzmli](https://github.com/wzmli) and [@papsti](https://github.com/papsti) at [@phac-nml-phrsd](https://github.com/phac-nml-phrsd) using  [`macpan2`](https://github.com/canmod/macpan2) modelling software. **This package is still in development.**

The goal of this package is to document the iterations of this model, so that they can be pulled into project-specific pipelines to produce modelling outputs. 

## Installation

You can install the development version of `EPACmodel` like so:

```{r eval = FALSE}
remotes::install_github("papsti/EPACmodel")
```

## Getting started guide

```{r}
library(EPACmodel)
```

### Set up model simulator

To work with a model, we need to load its simulator. A simulator in `macpan2` is an object that includes model structure (state names, flow expressions, etc.) along with a set of variable values (such as components of flow rates, initial states). 

This package includes several models whose simulators can quickly and easily be retrieved. Available models include:

```{r}
list_models()
```

To get this model's simulator, we simply call:

```{r}
model.name <- "two-age-groups"
model_simulator <- make_simulator(
  model.name = model.name
)
```

By default, `make_simulator()` will attach a set of default values for variables in the model (components of flow rates, initial states, etc.) to the model structure. It will also set a default number of time steps. You can see all default values easily as follows:

```{r}
# load default values (initial state, params, etc.)
default.values = get_default_values(model.name)
```

The definition of each default value is documented in each model's README (appended below in the [Available models](#available-models) section).

Any of these values can be changed by passing the optional `update.values` argument to `make_simulator()`. We recommend first loading the entire list of default values and extracting the specific value (list element) that you want to update, editing the numeric quantity as desired, and then passing the modified list element to `make_simulator()` via the `update.values` argument, to ensure that format of value is preserved to remain compatible with the model definition and `macpan2`:

```{r}
# get default initial state
default.state = default.values$state
print("default state:")
print(default.state)

# move some young susceptibles to the recovered class
new.state = default.state # copy over default value to preserve format
new.state["R_y"] = 1000 # modify specific elements
new.state["S_y"] = new.state["S_y"] - new.state["R_y"] # modify specific elements
print("new state:")
print(new.state)

# use updated state to make a new simulator 
new_model_simulator = make_simulator(
  model.name = model.name,
  updated.values = list(state = new.state)
)
```

### Simulate a model

To simulate a model, just use the `simulate()` function:

```{r}
sim_output = simulate(model_simulator)
```

Since the model simulator already has all required values attached, all required calculations can be performed to produce the simulation results. Simulation outputs may include several items, such as the count of individuals in each state (`value_type == "state"`) or the total inflow into each compartment (`value_type == "total inflow"`) at each time. 

For instance, here is the number of individuals in each state, stratified by the two age groups `y` (young) and `o` (old), at time 10:

```{r}
(sim_output
 |> dplyr::filter(value_type == 'state', time == 10)
)
```

The total inflow can be used to extract disease incidence over time:

```{r}
(sim_output
 |> dplyr::filter(
   stringr::str_detect(state_name, "^I"),
   value_type == 'total_inflow'
 )
 |> head()
)
```

Simulation outputs are documented in each model's README (appended below in the [Available models](#available-models) section). 

We can plot the results using standard data manipulation and plotting tools, like `dplyr` and `ggplot2`:

```{r sim-output-1, echo = FALSE}
tidy_output = function(output){
   age.lookup = tibble::tibble(
    age_label = c("y", "o"),
    age = c("young", "old")
  )
   sep.state = "_"
  
  (output
    # parse state names
     |> tidyr::separate(
       state_name,
       into = c("epi_state", "age_label"),
       sep = sep.state
     )
     |> dplyr::left_join(
       age.lookup,
       by = dplyr::join_by(age_label)
     )
     |> dplyr::select(-age_label)
     |> dplyr::filter(value_type == 'total_inflow', epi_state == 'I')
   )
}

plot_output <- function(df){
  (df
     |> ggplot2::ggplot()
     + ggplot2::geom_line(ggplot2::aes(x = time, y = value, colour = age),
                          linewidth = 1.25)
     + ggplot2::scale_y_continuous(
       labels = scales::label_number(scale_cut = scales::cut_short_scale())
     )
     + ggplot2::labs(
       x = "Days since start of outbreak",
       title = "Incidence over time by age group"
     )
     + ggplot2::theme_bw()
     + ggplot2::theme(
       axis.title.y = ggplot2::element_blank(),
       legend.position = c(1,1),
       legend.justification = c(1,1),
       legend.background = ggplot2::element_rect(fill = NA)
      )
  )
}

(sim_output
  # these are not package functions, defined just for this doc
 |> tidy_output()
 |> plot_output()
)
```

### Scenarios

By default, `make_simulator()` will initialize a base model. Some model definitions include optional scenarios on top of the base model, such as interventions modelled through time-varying model parameters. For instance, one could simulate a stay-at-home order by reducing the transmission rate on a given date by some amount. 

One can specify the `scenario.name` argument in `make_simulator()` to attach the required model structure to simulate a given scenario type. Scenario options and descriptions are catalogued in each model's README (appended below in the [Available models](#available-models) section).

Here we demonstrate the `two-age-groups` model with the `transmission-intervention` scenario. By default, this scenario reduces the transmission rate in each age group to 50% then 10% of its original value on days 30 and 40, respectively:

```{r}
values = get_default_values("two-age-groups")

values[
  c("intervention.day", "trans.factor.young", "trans.factor.old")
]
```

We specify that we want to use the `change-transmission` scenario in the call to `make_simulator()`:

```{r}
model_simulator <- make_simulator(
  model.name = "two-age-groups",
  scenario.name = "change-transmission"
)

sim_output = simulate(model_simulator)
```

```{r sim-output-2, echo = FALSE}
annotation = tibble::tibble(
  x = values$intervention.day,
  y = c(1.5e5, 7.5e4),
  label = paste0("Change transmission to\n",
                 values$trans.factor.young*100
                 ,"% original value")
)

(sim_output
 |> tidy_output()
 |> plot_output()
  # add annotations for interventions
  + ggplot2::geom_vline(
      data = annotation,
      mapping = ggplot2::aes(xintercept = x),
      linetype = "dashed",
      alpha = 0.5
    )
  + ggplot2::geom_label(
      data = annotation,
      mapping = ggplot2::aes(x = x, y = y, label = label),
      alpha = 0.5,
      size = 2.5
    )
)
```

## Available models {#available-models}

```{r, echo = FALSE}
model_readmes <- c()
for (mod in rev(list_models())){
  model_readmes <- c(model_readmes,
                     fs::path_package("models", mod, "README.Rmd",
                                      package = "EPACmodel"))
}
```

```{r, child = model_readmes, echo = FALSE}
```
