---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  # eval = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.align = "center",
  out.width = "75%"
)
```

# EPACmodel

<!-- badges: start -->
<!-- badges: end -->

This package implements the Early Pandemic Age-Structured Compartmental (EPAC) model developed by [Michael WZ Li](https://github.com/wzmli) and [Irena Papst](https://github.com/papsti) from the [Public Health Risk Sciences Division](https://github.com/phac-nml-phrsd) of the Public Health Agency of Canada using  [`macpan2`](https://github.com/canmod/macpan2) modelling software. 

The goal of this package is to catalogue, document, and version iterations of the Early Pandemic Age-structured Compartmental model so that they can be pulled easily into project-specific pipelines to produce modelling outputs. 

## Installation

Versioned releases of `EPACmodel` can be installed with

```{r eval = FALSE}
remotes::install_github("phac-nml-phrsd/EPACmodel@vx.y.z")
```

where `x.y.z` is the version number. Available version numbers can be found [here](https://github.com/phac-nml-phrsd/EPACmodel/releases).

The development version of `EPACmodel` can be installed with:

```{r eval = FALSE}
remotes::install_github("phac-nml-phrsd/EPACmodel")
```

## Getting started

A getting started guide can be found in `vignette("EPACmodel")`.

## For developers

Please start by reading the [developer notes](https://phac-nml-phrsd.github.io/EPACmodel/articles/dev).

