---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# EPAC model

<!-- badges: start -->
<!-- badges: end -->

This package implements the Early Pandemic Age-Structured Compartmental (EPAC) model developed by [@wzmli](https://github.com/wzmli) and [@papsti](https://github.com/papsti) at [@phac-nml-phrsd](https://github.com/phac-nml-phrsd) using  [`macpan2`](https://github.com/canmod/macpan2) modelling software. **This package is still in development.**

The goals of this package are to document the model, provide tools for simulating and calibrating the model, as well as visualizing model outputs.

## Installation

You can install the development version of EPACmodel like so:

``` r
# FILL THIS IN! HOW CAN PEOPLE INSTALL YOUR DEV PACKAGE?
```

## Getting started

```{r}
library(EPACmodel)
```

First, load a model. You can list the available model names here:

```{r}
list_models()
```

```{r}
model_name = "two-age-groups"
model = get_model(model_name)
```

This object encodes only the model _structure_. In order to simulate from this model, we need to attach parameters, as well as an initial state. 

The default parameters and initial state for each model can be loaded with:

```{r}
params = get_params(model_name)
state = get_state(model_name)
```

Once these vectors are loaded, you can modify manually in-session if you like before constructing the model simulator.

Once we have the model pieces, we need to put them together to construct a **simulator**, which is an object that we can use to numerically simulate the model:
