---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  # eval = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.align = "center",
  out.width = "75%"
)
```

# EPACmodel

<!-- badges: start -->
<!-- badges: end -->

This package implements the Early Pandemic Age-Structured Compartmental (EPAC) model developed by Michael WZ Li ([@wzmli](https://github.com/wzmli)) and Irena Papst ([@papsti](https://github.com/papsti)) from the Public Health Risk Sciences Division of the Public Health Agency of Canada ([@phac-nml-phrsd](https://github.com/phac-nml-phrsd))q using  [`macpan2`](https://github.com/canmod/macpan2) modelling software. 

The goal of this package is to document the iterations of this model, and package them so that they can easily be pulled into project-specific pipelines to produce modelling outputs. 

## Installation

The latest release of `EPACmodel` is [version 1.1.0](https://github.com/phac-nml-phrsd/EPACmodel/releases/tag/v1.1.0), and can be installed with:

```{r eval = FALSE}
remotes::install_github("phac-nml-phrsd/EPACmodel@v1.1.0")
```

You can install the development version of `EPACmodel` with:

```{r eval = FALSE}
remotes::install_github("phac-nml-phrsd/EPACmodel")
```
