---
title: "Introduction to EPACmodel"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to EPACmodel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(EPACmodel)
library(tibble)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(scales)
```

The goal of this package is to catalogue, document, and version iterations of the Early Pandemic Age-structured Compartmental model so that they can be pulled easily into project-specific pipelines to produce modelling outputs. 

## Set up model simulator

To work with a model, we need to set up its simulator. A simulator in `macpan2` is an object that includes model structure (state names, flow expressions, etc.) along with a set of variable values (such as components of flow rates, initial states) that ensure it is ready to produce simulation results. 

This package includes several models whose simulators can quickly and easily be retrieved (see `vignette("pkg-models")`). It can also work with locally defined models (see `vignette("local-models")`). 

Package model names can be listed with:

```{r}
list_models()
```

To get a model's simulator, we simply call:

```{r}
model.name <- "old-and-young"
model_simulator <- make_simulator(
  model.name = model.name
)
```

By default, `make_simulator()` will attach a set of default values for variables in the model (components of flow rates, initial states, etc.) to the model structure. It will also set a default number of time steps. You can see all default values easily as follows:

```{r}
# load default values (initial state, params, etc.)
default.values = get_default_values(model.name)
```

The definition of each default value is documented in `vignette("pkg-models")`.

Any of these values can be changed by passing the optional `values` argument to `make_simulator()`. We recommend first loading the entire list of default values and extracting the specific value (list element) that you want to update, editing the numeric quantity as desired, and then passing the modified list element to `make_simulator()` via the `values` argument, to ensure that format of value is preserved to remain compatible with the model definition and `macpan2`:

```{r}
# get default initial state
default.state = default.values$state
print("default state:")
print(default.state)

# move some young susceptibles to the recovered class
new.state = default.state # copy over default value to preserve format
new.state["R_y"] = 1000 # modify specific elements
new.state["S_y"] = new.state["S_y"] - new.state["R_y"] # modify specific elements
print("new state:")
print(new.state)

# use updated state to make a new simulator 
new_model_simulator = make_simulator(
  model.name = model.name,
  values = list(state = new.state)
)
```

(This approach works well if you want to update a parameter just once. However, if you want to repeatedly update some (or all) model parameters and re-run the simulator each time, this method would be costly as initializing a simulator takes a few seconds. Below, we discuss updating parameters after a model simulator is built, which is relatively fast.)

## Simulate a model

To simulate a model, just use the `simulate()` function:

```{r}
sim.output = simulate(model_simulator)
```

Since the model simulator already has all required values attached, all required calculations can be performed to produce the simulation results. 

For all models, simulation outputs are stored in a data frame with columns

```
time | state_name | value_type | value
```

The output `value_types` are

- `state`: the number of individuals in given state at a given time,
- `total_inflow`: the total inflow into a given compartment at a given time.

For instance, here is the number of individuals in each state, stratified by the two age groups `y` (young) and `o` (old), at time 10:

```{r}
(sim.output
 |> dplyr::filter(value_type == 'state', time == 10)
)
```

The total inflow into $I$ compartments can be used to extract disease incidence by age over time:

```{r}
(sim.output
 |> dplyr::filter(
   stringr::str_detect(state_name, "^I"),
   value_type == 'total_inflow'
 )
 |> head()
)
```

We can plot the results using standard data manipulation and plotting tools, like `dplyr` and `ggplot2`:

```{r sim-output-1, echo = FALSE}
tidy_output = function(output){
   age.lookup = tibble::tibble(
    age_label = c("y", "o"),
    age = c("young", "old")
  )
   sep.state = "_"
  
  (output
    # parse state names
     |> tidyr::separate(
       state_name,
       into = c("epi_state", "age_label"),
       sep = sep.state
     )
     |> dplyr::left_join(
       age.lookup,
       by = dplyr::join_by(age_label)
     )
     |> dplyr::select(-age_label)
     |> dplyr::filter(value_type == 'total_inflow', epi_state == 'I')
   )
}

plot_output <- function(df, ylim = NULL){
  (df
     |> ggplot2::ggplot()
     + ggplot2::geom_line(ggplot2::aes(x = time, y = value, colour = age),
                          linewidth = 1.25)
     + ggplot2::scale_y_continuous(
       limits = ylim,
       labels = scales::label_number(scale_cut = scales::cut_short_scale())
     )
     + ggplot2::labs(
       x = "Days since start of outbreak",
       title = "Disease incidence over time",
       colour = "Age group"
     )
     + ggplot2::theme_bw(
       base_size = 16
     )
     + ggplot2::theme(
       axis.title.y = ggplot2::element_blank(),
       legend.position = "inside",
       legend.position.inside = c(0.95, 0.95),
       legend.justification = c("right", "top"),
       legend.background = ggplot2::element_rect(fill = NA),
       legend.key = ggplot2::element_rect(fill = NA)
      )
  )
}

(sim.output
  # these are not package functions, defined just for this doc
 |> tidy_output()
 |> plot_output()
)
```

We could also update model parameters before the simulation using the `values` argument.[^1] Here, it's important to pass the **complete** list of model values (with any desired updates).[^2] As a result, we suggest starting by loading a complete list of the model's values with `get_default_values()`, updating the desired values, and then passing this list to `simulate()` as below. Note that the `time.steps` parameter cannot be updated after the model simulator is built and must instead be modified using the `values` argument of `make_simulator()`.

[^1]: Currently only supported for the `hosp` model.
[^2]: As opposed to with `make_simulator()` where we could pass only a partial parameter list and use defaults for the remaining parameters. This requirement could be relaxed in future versions.

```{r}
model.name2 <- "hosp"
model_simulator2 <- make_simulator(model.name2, values = list(time.steps = 365))
```

```{r echo=FALSE}
(simulate(model_simulator2)
  # these are (non-exported) package functions
 |> EPACmodel:::tidy_output()
 |> EPACmodel:::aggregate_across_age_groups()
 |> EPACmodel:::aggregate_across_epi_subcats()
 |> EPACmodel:::plot_output()
)
```

```{r}
values2 = get_default_values(model.name2)
values2$transmissibility <- values2$transmissibility*(1-0.2) # reduce transmissibility by 20%
```

```{r echo=FALSE}
(simulate(model_simulator2, values2)
  # these are (non-exported) package functions
 |> EPACmodel:::tidy_output()
 |> EPACmodel:::aggregate_across_age_groups()
 |> EPACmodel:::aggregate_across_epi_subcats()
 |> EPACmodel:::plot_output()
)
```

## Scenarios

By default, `make_simulator()` will initialize a base model. Some model definitions include optional scenarios on top of the base model, such as interventions modelled through time-varying model parameters. For instance, one could simulate a stay-at-home order by reducing the transmission rate on a given date by some amount. 

One can specify the `scenario.name` argument in `make_simulator()` to attach the required model structure to simulate a given scenario type. Scenario options and descriptions are catalogued in `vignette("pkg-models")`.

Here we demonstrate the `old-and-young` model with the `transmission-intervention` scenario. By default, this scenario reduces the transmission rate in each age group to 50% then 10% of its original value on days 30 and 40, respectively:

```{r}
values = get_default_values("old-and-young")

values[
  c("intervention.day", "trans.factor.young", "trans.factor.old")
]
```

We specify that we want to use the `change-transmission` scenario in the call to `make_simulator()`:

```{r}
model_simulator <- make_simulator(
  model.name = "old-and-young",
  scenario.name = "change-transmission"
)

sim.output = simulate(model_simulator)
```

```{r sim-output-2, echo = FALSE}
annotation = tibble::tibble(
  x = values$intervention.day,
  y = c(2.45e5, 2e5),
  label = paste0("Change transmission to\n",
                 values$trans.factor.young*100
                 ,"% original value")
)

(sim.output
 |> tidy_output()
 |> plot_output(
   ylim = c(0, 2.6e5)
 )
  # add annotations for interventions
  + ggplot2::geom_vline(
      data = annotation,
      mapping = ggplot2::aes(xintercept = x),
      linewidth = 1.1,
      alpha = 0.25
    )
  + ggplot2::geom_label(
      data = annotation,
      mapping = ggplot2::aes(x = x, y = y, label = label)
    )
)
```
